---
title: "AML Somatic Mutation Landscape — TCGA-LAML Cohort"
author: "Maria Antonietta Diana"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 10, fig.height = 6)
```

## Dataset

This analysis examines the somatic mutation landscape of **Acute Myeloid Leukaemia (AML)** using open-access data from The Cancer Genome Atlas (**TCGA-LAML** cohort, n=153 patients).

Somatic mutation calls were downloaded from the GDC Data Portal (masked somatic mutations, Mutect2 ensemble calls, hg38). Individual per-sample MAF files were merged using `scripts/part2_aml_cohort/03_merge_maf_files.py`.

---

## Setup

```{r libraries}
library(maftools)
library(tidyverse)
library(RColorBrewer)
```

```{r load-maf}
maf <- read.maf("../data/processed/TCGA-LAML-merged.maf")
```

---

## Cohort Summary

```{r summary}
getSampleSummary(maf) %>% head(10)
```

```{r gene-summary}
getGeneSummary(maf) %>% head(20)
```

```{r plot-summary, fig.height=8}
plotmafSummary(maf,
               rmOutlier = TRUE,
               addStat   = "median",
               dashboard = TRUE,
               titvRaw   = FALSE)
```

---

## Oncoprint

The oncoprint shows the most frequently mutated genes across the cohort. Each column is a patient, each row is a gene.

```{r oncoprint, fig.height=8, fig.width=14}
oncoplot(maf        = maf,
         top        = 20,
         removeNonMutated = TRUE)
```

---

## Most Frequently Mutated AML Genes

```{r top-genes, fig.height=5}
gene_summary <- getGeneSummary(maf)

gene_summary %>%
  arrange(desc(AlteredSamples)) %>%
  head(15) %>%
  mutate(Gene = fct_reorder(Hugo_Symbol, AlteredSamples)) %>%
  ggplot(aes(x = Gene, y = AlteredSamples / getSampleSummary(maf) %>% nrow() * 100)) +
  geom_col(fill = "#2196F3", width = 0.7) +
  coord_flip() +
  labs(title = "Top 15 Mutated Genes in TCGA-LAML",
       x = NULL,
       y = "% Patients Mutated") +
  theme_minimal(base_size = 13)
```

---

## Lollipop Plots — Key AML Drivers

```{r lollipop-flt3, fig.height=4}
lollipopPlot(maf     = maf,
             gene    = "FLT3",
             AACol   = "HGVSp_Short",
             showMutationRate = TRUE,
             labelPos = "all")
```

```{r lollipop-npm1, fig.height=4}
lollipopPlot(maf     = maf,
             gene    = "NPM1",
             AACol   = "HGVSp_Short",
             showMutationRate = TRUE)
```

```{r lollipop-dnmt3a, fig.height=4}
lollipopPlot(maf     = maf,
             gene    = "DNMT3A",
             AACol   = "HGVSp_Short",
             showMutationRate = TRUE)
```

---

## Mutational Signatures

```{r signatures, fig.height=6, eval=FALSE}
maf_tnm <- trinucleotideMatrix(
  maf            = maf,
  ref_genome     = "BSgenome.Hsapiens.UCSC.hg38",
  ignoreChr      = "chrM",
  useSyn         = TRUE
)
```

```{r sigs-plot, eval=FALSE}
plotSignatures(maf_tnm,
               title_size = 1.2)
```

---

## Co-occurrence and Mutual Exclusivity

```{r cooccurrence, fig.height=8}
somaticInteractions(maf     = maf,
                    top     = 20,
                    pvalue  = c(0.05, 0.1))
```

---

## VAF Distribution

```{r vaf, fig.height=5}
plotVaf(maf = maf, vafCol = "t_alt_count")
```

---

## Interpretation

```{r interpretation, results='asis'}
gs <- getGeneSummary(maf)
n_patients  <- getSampleSummary(maf) %>% nrow()
n_mutations <- nrow(maf@data)

cat(paste0(
  "### Key Findings\n\n",
  "- **Cohort size**: ", n_patients, " AML patients\n",
  "- **Total somatic mutations**: ", n_mutations, "\n",
  "- **Median mutations per patient**: ",
  median(getSampleSummary(maf)$total), "\n\n"
))

cat("### Top mutated genes and their biological significance\n\n")

top_genes <- gs %>%
  arrange(desc(AlteredSamples)) %>%
  head(10)

annotations <- list(
  FLT3   = "Class I mutation — activates proliferation signalling. ~25% of AML. Targetable with midostaurin/gilteritinib.",
  NPM1   = "Class II mutation — causes nucleophosmin mislocalisation. ~30% of AML. Favourable prognosis when FLT3-ITD negative.",
  DNMT3A = "Epigenetic regulator — DNA methyltransferase. ~20% of AML. Often founding clonal event.",
  IDH1   = "Produces oncometabolite 2-HG, blocks differentiation. Targetable with ivosidenib.",
  IDH2   = "Same pathway as IDH1. Targetable with enasidenib.",
  TP53   = "Tumour suppressor — loss of function. Associated with complex karyotype and poor prognosis.",
  RUNX1  = "Transcription factor — core binding factor AML.",
  TET2   = "Epigenetic regulator — DNA demethylation. Associated with clonal haematopoiesis.",
  NRAS   = "RAS signalling — Class I proliferation mutation.",
  CEBPA  = "Transcription factor — biallelic mutations confer favourable prognosis."
)

for (i in seq_len(nrow(top_genes))) {
  gene <- top_genes$Hugo_Symbol[i]
  pct  <- round(top_genes$AlteredSamples[i] / n_patients * 100, 1)
  ann  <- annotations[[gene]]
  if (!is.null(ann)) {
    cat(paste0("- **", gene, "** (", pct, "% of patients): ", ann, "\n"))
  }
}
```